---
title: "Universal covariates"
output: html_document
editor_options: 
  chunk_output_type: console
---

Note: as the coordinates of camera traps are sensitive data, the datasets needed to run this code are not publicly available. Complete data are available to qualified researchers from the Wildlife Chief Scientist of the Washington Department of Fish and Wildlife by contacting (360)-902-2515. This information is provided for clarity on decisions we made for covariate cleaning and use.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(camtrapR) 
library(tidyverse)
library(lubridate)
library(sf)
library(raster)
library(sp)
library(unmarked)
library(here)
select <- dplyr::select
```


## Station data
```{r}
stations_ok <- read_csv(here("data", "OK_Camera_Stations_18-19.csv")) %>% rename(Problem2_to = Problem2_from...10, Problem2_from = Problem2_from...9)
stations_ne <- read_csv(here("data","NE_Camera_Stations_18-19_v2.csv"))
stations <- rbind(stations_ne, stations_ok)
stations <- stations %>% mutate(CameraLocation=paste(Cell_ID, Camera_ID, sep="_")) %>% relocate(CameraLocation, .after="Camera_ID") %>% mutate(Cell_ID=recode_factor(Cell_ID, "NE2048a" = "NE2048", "NE2048b" = "NE2048", "NE5345a" = "NE5345", "NE5345b" = "NE5345"))
stations_1920 <- read_csv(here("data","All_Camera_Stations_19-20.csv"))
stations_1920$Notes <- NA
stations_2021 <- read_csv(here("data","All_Camera_Stations_20-21.csv"))
stations_2021$Notes <- NA
stations <- rbind(stations, stations_1920, stations_2021)
stations_temp <- stations %>%
  transmute(
    Cell_ID = as.factor(as.character(Cell_ID)),
    Camera_ID = as.character(Camera_ID), 
    UTM_X = Longitude,
    UTM_Y = Latitude,
    Setup_date = as.character(mdy(Setup_date)),
    Retrieval_date = as.character(mdy(Retrieval_date)),
    Problem1_from = as.character(mdy(Problem1_from)),
    Problem1_to = as.character(mdy(Problem1_to)),
    Problem2_from = as.character(mdy(Problem2_from)),
    Problem2_to = as.character(mdy(Problem2_to))
  ) 
stations2 <-  stations_temp %>% 
  filter(Cell_ID != "OK3667" | Camera_ID != 92 | !is.na(Problem1_from)) %>%
  dplyr::select(-Problem2_from, -Problem2_to)

# removing problem cameras
a <- which(stations2$Cell_ID == "OK2749" & stations2$Setup_date == "2018-07-08")
b <- which(stations2$Cell_ID == "OK7658" & stations2$Setup_date == "2018-08-22")
stations2 <- stations2[-c(a,b),]

c <- which(stations2$Cell_ID == "OK1474" & stations2$Setup_date == "2018-10-23")
stations2 <- stations2[-c,] %>%
  filter(Cell_ID != "NE2897", Cell_ID != "NE2899") 

stations2$CameraLocation <- paste(stations2$Cell_ID, stations2$Camera_ID, sep="_")
stations2$Year <- ifelse(year(stations2$Setup_date)==2018, 2018, ifelse(year(stations2$Setup_date)==2019, 2019, 2020))
stations2 <- stations2 %>% arrange(Year, CameraLocation)
```


## Extracting covariates

- double checking since Sarah B. had site covariates retained in her models: only issue could be different dataset versions
```{r, include = FALSE}
NE <-raster(here("data", "NE_DEM.img"))
NE@crs
OK <- raster(here("data", "MW_DEM.img"))
OK@crs
DEM <- merge(NE, OK)

##Camera Station Locations
cams_1819_NE <- read.csv(here("data", "NE_Camera_Stations_18-19_v2.csv")) %>%
  st_as_sf(crs = 4326, coords = c('Longitude', 'Latitude')) %>%
  mutate(Year = 2018)
cams_1819_OK <- read.csv(here("data", "OK_Camera_Stations_18-19.csv")) %>%
  st_as_sf(crs = 4326, coords = c('Longitude', 'Latitude')) %>%
  mutate(Year = 2018) %>%
  rename("Problem2_to" = "Problem2_from.1")
cams.18 <- rbind(cams_1819_NE, cams_1819_OK) %>%
  rename("Set_date"="Setup_date", "Pull_date"="Retrieval_date", "Problem1_f"="Problem1_from", "Problem1_t"="Problem1_to", "Problem2_f"="Problem2_from", "Problem2_t"="Problem2_to") %>%
  mutate(CameraLocation = paste(Cell_ID, Camera_ID, sep="_")) %>%
  relocate(CameraLocation, .after=Camera_ID) %>%
  select(-Notes)
cams.19 <- read_csv(here("data","All_Camera_Stations_19-20.csv")) %>%
  st_as_sf(crs = 4326, coords = c('Longitude', 'Latitude')) %>%
  mutate(Year = 2019) %>%
  rename("Set_date"="Setup_date", "Pull_date"="Retrieval_date", "Problem1_f"="Problem1_from", "Problem1_t"="Problem1_to", "Problem2_f"="Problem2_from", "Problem2_t"="Problem2_to")
cams.20 <- read_csv(here("data","All_Camera_Stations_20-21.csv")) %>%
  st_as_sf(crs = 4326, coords = c('Longitude', 'Latitude')) %>%
  mutate(Year = 2020) %>%
  rename("Set_date"="Setup_date", "Pull_date"="Retrieval_date", "Problem1_f"="Problem1_from", "Problem1_t"="Problem1_to", "Problem2_f"="Problem2_from", "Problem2_t"="Problem2_to")

covs <- rbind(cams.18, cams.19, cams.20) %>%
  dplyr::select(Cell_ID, Camera_ID, Year, geometry) %>%
  st_transform("+proj=lcc +lat_0=45.3333333333333 +lon_0=-120.5 +lat_1=45.8333333333333 +lat_2=47.3333333333333 +x_0=500000 +y_0=0
+ellps=GRS80 +units=m +no_defs") # to match DEM projection, not strictly necessary

# Extract the elevation (in meters) of each camera location
covs <- covs %>%
  mutate(elev = extract(DEM, covs))

# Landcover database
nlcd <- raster(here("data", "nlcd_wa_utm10.tif"))

# Extract the national land cover type at each camera location
covs <- covs %>%
  mutate(landcover = extract(nlcd, covs)) # automatically transforms covs to crs of nlcd

# distance to major roads from WADNR 
roads_ne <- st_read(here("data",'WADNR_Active_Roads_NE.shp'))
roads_ok <- st_read(here("data","WADNR_Active_Roads_OK.shp"))
roads <- rbind(roads_ne, roads_ok) # just taking distance from this I believe
roads <- roads %>% st_transform("+proj=lcc +lat_1=45.83333333333334 +lat_2=47.33333333333334
                                   +lat_0=45.33333333333334 +lon_0=-120.5 +x_0=500000 +y_0=0 +ellps=GRS80
                                   +units=m +no_defs")
closestRoad <- st_nearest_feature(covs, roads)
covs.2 <- covs %>% mutate(wadnr = NA) # distance is in meters 

for(i in 1:nrow(covs.2)) {
  covs.2$wadnr[i] = st_distance(covs.2$geometry[i], roads$geometry[closestRoad[i]])
} 

# other covariates 
cam.data <- read.csv(here("data","Camera_Station18-21_Covariates_2022-04-06.csv")) %>%
  mutate(Habitat_Type = as.factor(Habitat_Type), .keep = "unused") %>%
  mutate(Distance_to_Road = Distance_Focal_Point, .keep = "unused") %>% # distance to linear feature
  mutate(Cam_Height = Height_frm_grnd, .keep = "unused") %>%
  mutate(Road_Type = as.factor(Monitoring), .keep = "unused") %>%
  mutate(Canopy = Canopy_Cov, .keep = "unused") %>%
  mutate(Land_Mgnt = recode_factor(Land_Mgnt, "Federal" = "Public",
                            "State" = "Public",
                            "Private" = "Private"), .keep = "unused") %>%
  mutate(Year = ifelse(Year == "Year1", 2018, ifelse(Year == "Year2", 2019, 2020)))
  
# Add camera covariate data to covariates
# Fixes for renaming cells
covs.2$Cell_ID[covs.2$Cell_ID %in% c("NE2048a", "NE2048b")] <- "NE2048"
covs.2$Cell_ID[covs.2$Cell_ID %in% c("NE5345a", "NE5345b")] <- "NE5345"
covs.3 <- left_join(covs.2, cam.data, by=c("Cell_ID", "Year", "Camera_ID")) 

# Duplicates
duplicated_sites <- covs.3$Cell_ID[which(duplicated(covs.3$Cell_ID))]
covs.3 <- covs.3[-which(duplicated(covs.3$CameraLocation) & covs.3$Cell_ID=="NE5920"),]
```


### finalize site covariates/camera stations
```{r}
sitecovs <- covs.3 %>% arrange(Cell_ID) %>% dplyr::select("Cell_ID", "Camera_ID", "Year", "landcover", "Habitat_Type", "Road_Type", "Land_Mgnt", "elev", "Cam_Height", "Distance_to_Road", 'wadnr')%>% 
  mutate(Habitat_Type = recode_factor(Habitat_Type, "Mixed conifer" = "mixed conifer")) %>%
  mutate(Road_Type = recode_factor(Road_Type, "Decommissioned road" = "road", "Dirt road" = "road", "Game trail" = "trail", "Closed road" = "road", "Trail" = "trail")) %>%
  mutate(conifer = recode_factor(Habitat_Type, "mixed conifer" = "conifer", "Burned" = "other", "Grassland" = "other",  "Shrub-steppe" = "other", "Agriculture" = "other", "Riparian" = "other")) %>% 
  mutate(scale.elev = (elev - mean(elev)) / sd(elev)) %>%
  mutate(scale.dtr = (Distance_to_Road - mean(Distance_to_Road, na.rm = T))/sd(Distance_to_Road, na.rm = T))%>%
  mutate(scale.cheight = (Cam_Height -mean(Cam_Height, na.rm = T))/sd(Cam_Height, na.rm = T)) %>%
  mutate(scale.wadnr = (wadnr-mean(wadnr, na.rm = T))/sd(wadnr, na.rm = T)) %>%
  mutate(landmg2 = recode_factor(Land_Mgnt, "public" = "public",
                                 "private" = "private")) %>%
  # removing sites with missing covariates
  drop_na(conifer, landmg2, scale.elev, scale.wadnr, Road_Type, scale.cheight, scale.dtr)

sitecovs$conifer[which(sitecovs$conifer=="")] <- "other"
sitecovs$conifer <- droplevels(sitecovs$conifer)
sitecovs$CameraLocation <- paste(sitecovs$Cell_ID, sitecovs$Camera_ID, sep="_")

# only ones which are in stations2
sitecovs.2 <- semi_join(sitecovs, stations2, by = c("Cell_ID", "Camera_ID", "Year")) 

# removing duplicates/missing sitecovs
sitecovs.2 %>%
  group_by(Cell_ID, Camera_ID, Year) %>%
  mutate(how.many = n()) %>%
  distinct(Cell_ID, Camera_ID, Year, how.many) %>%
  filter(how.many > 1)

issues <- which(sitecovs.2$CameraLocation %in% c("OK2749_59", "OK3667_92", "OK7658_43") & duplicated(sitecovs.2$CameraLocation)) 
issues <- c(issues, which(sitecovs.2$Cell_ID=="OK1790" & sitecovs.2$Year == 2019))

sitecovs.2 <- sitecovs.2[-issues,]
missing <- anti_join(stations2, sitecovs.2, by = c("Cell_ID", "Camera_ID", "Year"))

stations.2 <- semi_join(stations2, sitecovs.2, by = c("Cell_ID", "Camera_ID", "Year"))

write_csv(sitecovs.2, here("data",'tidy_covs.csv'))
sitecovs.2 <- read_csv(here("data","tidy_covs.csv"))

write_csv(stations.2, here("data",'stationsTbl.csv'))
stations.2 <- read_csv(here("data",'stationsTbl.csv'))

stations.2.scrubbed <- stations.2 %>%
  dplyr::select(-UTM_X, -UTM_Y)

write_csv(stations.2.scrubbed, here("data-tidy", "stationsTbl.csv"))
```


## Observational covariates

### smoke-based pm 2.5
```{r}
# data
smoke.2018 <- read_csv(here("data", "smoke_pm_2_5_2018.csv")) %>%
  mutate(Year = 2018)
smoke.2019 <- read_csv(here("data", "smoke_pm_2_5_2019.csv")) %>% 
  mutate(Camera_ID = as.character(Camera_ID)) %>%
  mutate(Year = 2019)
smoke.2020 <- read_csv(here("data", "smoke_pm_2_5_2020.csv")) %>%
  mutate(Year = 2020)

# rename to not have year
smoke.2018.2 <- smoke.2018 %>%
  rename_with(~ str_sub(.x, 6, 10), starts_with("20"))
smoke.2019.2 <- smoke.2019 %>%
  rename_with(~ str_sub(.x, 6, 10), starts_with("20"))
smoke.2020.2 <- smoke.2020 %>%
  rename_with(~ str_sub(.x, 6, 10), starts_with("20"))

# select date range + add blank columns for missing dates
smoke.2018.dates <- smoke.2018.2 %>%
  select("07-01":"10-30", Camera_ID, Year, .geo) %>%
  mutate(`10-31` = NA) %>% relocate(`10-31`, .after = "10-30") %>%
  mutate(`10-10` = NA) %>% relocate(`10-10`, .after = "10-09")

smoke.2019.dates <- smoke.2019.2 %>%
  select("07-01":"10-30", Camera_ID, Year, .geo) %>%
  mutate(`07-05` = NA) %>% relocate(`07-05`, .after = "07-04") %>%
  mutate(`07-06` = NA) %>% relocate(`07-06`, .after = "07-05") %>%
  mutate(`07-07` = NA) %>% relocate(`07-07`, .after = "07-06") %>%
  mutate(`07-12` = NA) %>% relocate(`07-12`, .after = "07-11") %>%
  mutate(`07-17` = NA) %>% relocate(`07-17`, .after = "07-16") %>%
  mutate(`08-02` = NA) %>% relocate(`08-02`, .after = "08-01") %>%
  mutate(`09-11` = NA) %>% relocate(`09-11`, .after = "09-10") %>%
  mutate(`09-12` = NA) %>% relocate(`09-12`, .after = "09-11") %>%
  mutate(`10-18` = NA) %>% relocate(`10-17`, .after = "10-16") %>%
  mutate(`10-31` = NA) %>% relocate(`10-31`, .after = "10-30")

smoke.2020.dates <- smoke.2020.2 %>%
  dplyr::select("07-01":"10-31", Camera_ID, Year, .geo) %>%
  mutate(`07-23` = NA) %>% relocate(`07-23`, .after = "07-22") %>%
  mutate(`07-29` = NA) %>% relocate(`07-29`, .after = "07-28") %>%
  mutate(`07-30` = NA) %>% relocate(`07-30`, .after = "07-29")

# together
smoke.all <- rbind(smoke.2018.dates, smoke.2019.dates, smoke.2020.dates)

# bind to stations2 to compare location data for duplicated entries
smoke.weird <- left_join(stations.2, smoke.all, by = c("Camera_ID", "Year")) %>%
  filter((Camera_ID == 59 & Year == 2020) | 
           Camera_ID == 89 & Year == 2020 |
           Camera_ID == 96 & Year == 2019) %>%
  arrange(Camera_ID)

smoke.okay <- left_join(stations.2, smoke.all, by = c("Camera_ID", "Year")) %>%
  anti_join(smoke.weird, by = c("Camera_ID", "Year"))

smoke.weird.2 <- smoke.weird %>%
  select(Year, Cell_ID, Camera_ID, UTM_X, UTM_Y, .geo)
# 1, 4, 5, 8, 9, 12 match by location

smoke.weird.keep <- smoke.weird[c(1, 4, 5, 8, 9, 12),]

smoke.fixed <- rbind(smoke.okay, smoke.weird.keep)

# order according to sitecovs, which is what the other ones are ordered by
smoke.ordered <- left_join(sitecovs.2, smoke.fixed, by = c("Cell_ID", "Camera_ID", "Year"))

smoke.final <- smoke.ordered %>%
  as_tibble() %>%
  select("07-01":"10-31") %>%
  as.matrix()

# describe before removing outliers
smoke.megalist <- smoke.final %>% c() %>% unlist()
summary(smoke.megalist)
m <- mean(smoke.megalist, na.rm = T) #7.432
s <- sd(smoke.megalist, na.rm = T) # 41.521

cutoff <- quantile(smoke.megalist, 0.999, na.rm = T) # 267.905
smoke.final.2 <- as.matrix(smoke.final)
remove <- which(smoke.final.2 > 267.7699) # not row,col format

smoke.final.cutoff <- smoke.final.2
smoke.final.cutoff[remove] <- NA

# describe after removing outliers (modeling data set)
smoke.megalist.2 <- smoke.final.cutoff %>% c() %>% unlist()
summary(smoke.megalist.2)
m <- mean(smoke.megalist.2, na.rm = T) # 6.597
s <- sd(smoke.megalist.2, na.rm = T) # 21.247

smoke.scaled <- smoke.final.cutoff %>%
  as_tibble() %>%
  mutate(across(where(is.numeric), ~ (.x - m)/s)) %>%
  as.matrix()

# save
write.table(smoke.scaled, here("data", "scaled_smoke_matrix.csv"))
write.table(smoke.final, here("data", "raw_smoke_matrix.csv"))
```


### adding temperature locations

```{r new_temperature_processing}
temptemp <- read_csv(here("data","WPPP_weather_land_canopy_roads_sunrise_PM2.5.csv")) 
study_interval1 <- interval(ymd("2018-07-01"), ymd("2018-10-31")) 
study_interval2 <- interval(ymd("2019-07-01"), ymd("2019-10-31")) 
study_interval3 <- interval(ymd("2020-07-01"), ymd("2020-10-31")) 

temp1 <- temptemp %>% 
  mutate(Obs_Date = Observation_Date, Air_Temperature_K=Daily_Mean_Air_Temperature.K., Precipitation_mm=Daily_Accumulated_Precipitation.mm.) %>%
  mutate(CameraLocation=Camera_Location) %>%
  filter(Obs_Date %within% study_interval1 | Obs_Date %within% study_interval2 | Obs_Date %within% study_interval3) %>%
  dplyr::select(Cell_ID, Camera_ID, CameraLocation, Year, Obs_Date, Air_Temperature_K, Precipitation_mm) %>%
  mutate(temp.k.scaled = scale(Air_Temperature_K)) %>%
  mutate(precip.mm.scaled = scale(Precipitation_mm)) %>%
  mutate(Year = ifelse(Year == "Year1", 2018, ifelse(Year == "Year2", 2019, 2020)))

temp_matrix <- matrix(nrow = nrow(sitecovs.2), ncol = 123)
temp_matrix.scaled <- matrix(nrow = nrow(sitecovs.2), ncol = 123)
for(i in 1:nrow(temp_matrix)){ 
  for (j in 1:(ncol(temp_matrix))) { 
    temp_date <- c(ymd("2018-06-30") + days(j), ymd("2019-06-30") + days(j), ymd("2020-06-30") + days(j))
    b = which(temp1$Obs_Date %in% temp_date & temp1$Cell_ID == sitecovs.2$Cell_ID[i] & temp1$Year == sitecovs.2$Year[i])
    b = min(b) 
    if (is.finite(b)) { 
      temp_matrix[i,j] = temp1$Air_Temperature_K[b]
      temp_matrix.scaled[i,j] = temp1$temp.k.scaled[b]
    } 
  }
} 

temp_missing_rows <- unique(which(is.na(temp_matrix), arr.ind=T)[,1])

write.table(temp_matrix, here("data", "unscaled_temp_matrix.csv"))
write.table(temp_matrix.scaled, here("data",'scaled_temp_obs_cov.csv'))
```

### summarizing temperature and pm 2.5 conditions
temperature:
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
-12.431   8.696  14.392  13.395  18.598  30.464    2876  
super simplistic summary: just all observations from all camera traps, days, and years 

pm 2.5:
   Min.  1st Qu.   Median     Mean  3rd Qu.     Max.  
   0.000    0.017    0.199    7.432    2.181 4213.255     1856 
   sd = 41.521
annual standard: 12 mug/m3, happened 11.7% of camera-days
24 hr standard: 35 mug/m3, happened 5.5% of camera-days

temperature and smoke were correlated at r = 0.099 (very not correlated)

temperature above 17-24 C in 3.8 to 33.8% of CT days
```{r}
temp_unscaled <- read.table(here("data", "unscaled_temp_matrix.csv"))
temp.megalist <- temp_unscaled %>%
  c() %>%
  unlist() %>%
  -273.15 
summary(temp.megalist)
sum(temp.megalist > 24, na.rm = T)/(length(temp.megalist)-sum(is.na(temp.megalist)))
sum(temp.megalist > 17, na.rm = T)/(length(temp.megalist)-sum(is.na(temp.megalist)))

smoke_unscaled <- read.table(here("data", "raw_smoke_matrix.csv"))
s <- smoke.final %>% as.matrix() %>% sd(na.rm = T)
pm2.5.megalist <- smoke_unscaled %>%
  as.matrix() %>%
  c() 
summary(pm2.5.megalist) # wild outliers: remove?? (yes, did)
hist(pm2.5.megalist)  
sum(pm2.5.megalist > 12, na.rm = T)/(length(pm2.5.megalist)-sum(is.na(pm2.5.megalist)))
sum(pm2.5.megalist > 35, na.rm = T)/(length(pm2.5.megalist)-sum(is.na(pm2.5.megalist)))
sum(pm2.5.megalist > 100, na.rm = T)/(length(pm2.5.megalist)-sum(is.na(pm2.5.megalist)))

temp <- temp_unscaled %>% as.matrix()
smoke <- smoke_unscaled  %>% as.matrix()
temp[which(is.na(smoke))] <- NA 
smoke[which(is.na(temp))] <- NA 
cor(na.omit(as.vector(temp)), na.omit(as.vector(smoke)))
```


### days
```{r}
#  number of camera sites
R <- nrow(sitecovs.2)
#  total number of days in season of interest
D <- 123

date.vector <- rep(c(1:D))
date.vector.scaled <- (date.vector - mean(date.vector))/sd(date.vector)
days <- matrix(rep(c(1:D), R), ncol = D, nrow = R, byrow = T)
scale.days <- matrix(rep(date.vector.scaled, R), ncol = D, byrow = T)

write.table(scale.days, here("data",'scaledDaysObsCov.csv'))
scale.days <- read.table(here("data",'scaledDaysObsCov.csv'))
```

### among datasets
```{r}
sitecovs.2 <- read_csv(here("data","tidy_covs.csv")) %>% select(-geometry)
temp_scaled.2 <- read.table(here("data",'scaled_temp_obs_cov.csv')) %>% as.matrix()
days_scaled.2 <- read.table(here("data",'scaledDaysObsCov.csv')) %>% as.matrix()
smoke_scaled.2 <- read.table(here("data", "scaled_smoke_matrix.csv")) %>% as.matrix()
smoke_unscaled <- read.table(here("data", "raw_smoke_matrix.csv")) %>% as.matrix()

# aligning data: need to all have the same number of sites as sitecovs
nrow(sitecovs.2)
nrow(temp_scaled.2)
nrow(days_scaled.2)
nrow(smoke_scaled.2)

detection_history_order <- order(sitecovs.2$Year, sitecovs.2$CameraLocation) # creating new order
sitecovs.2 <- sitecovs.2[detection_history_order,]
smoke_scaled.2 <- smoke_scaled.2[detection_history_order,] # these were ordered by sitecovs originally
temp_scaled.2 <- temp_scaled.2[detection_history_order,]
smoke_unscaled <- smoke_unscaled[detection_history_order,]

## make all obs covs have the same NAs to keep sample size the same during modeling
temp_scaled.2[which(is.na(smoke_scaled.2))] <- NA 
smoke_scaled.2[which(is.na(temp_scaled.2))] <- NA
days_scaled.2[which(is.na(temp_scaled.2))] <- NA
# skipping for unscaled smoke since i'm using it for a visual only

# write final bits
write_csv(sitecovs.2, here("data-tidy", "covs-site.csv"))
write.table(smoke_scaled.2, here("data-tidy", "covs-obs-smoke-scaled.csv"))
write.table(days_scaled.2, here("data-tidy", "covs-obs-day-scaled.csv"))
write.table(temp_scaled.2, here("data-tidy", "covs-obs-temp-scaled.csv"))
write.table(smoke_unscaled, here("data", "unscaled_smoke_matrix2.csv"))
```

